---
title: ""
author: "Andrew Tolonen"
format: html
editor: visual
toc: TRUE
code-fold: TRUE
embed-resources: TRUE
---

# Introduction

This notebook describes methods to visualize and filter flow cytometry data in .FCS files using the R flowCore package. Data was acquired using a Beckman CytoFlex instrument. Here, the samples are SYBR-stained bacterial cells. The goal is to accurately enumerate the number of bacteria in a defined volume.

## Background

FCS files contain all acquired data. However, running Cytoflex requires 3 user-defined settings that affect what data is acquired:

-   Threshold = min signal for event, should set min signal manually.

-   Gain (hand signal on Cytoflex software). Note, the abort rate goes up if gain is too high.

-   Width = Set a width threshold, but other users stay with the default.

Also, the software asks the user to calculate number of events to save. This can be total events or total time. We defined a 20-30 sec of acquisition period with a flow rate of 10 ul per min. Thus, the sample volume was 3.33 ul.

# Setup

Libraries

```{r Libraries, message = F}

# libraries to include
library(knitr);
library(tidyverse);
library(readxl);
library(readr);
library(gridExtra);
library(ggpubr);
library(glue);

library(flowCore);
library(ggcyto);
library(lattice);

```

Flags/Config

```{r flags and config}

rm(list = ls());

mytheme = theme(axis.text.x = element_text(size = 8), 
                axis.text.y = element_text(size = 8), 
                axis.title.x = element_blank(), 
                axis.title.y = element_text(size = 8),
                strip.text.x = element_text(size = 8),
                strip.text.y = element_text(size = 8),
                legend.position = "none", 
                aspect.ratio = 0.25,
                plot.title = element_text(size = 8),
               );

# dilution factor of sample
dilution_factor = 4;

```

File IO

Note: I am not truncating the data even though some channels exceed the max \$PnR value

```{r File IO}

path_in = "/home/tolonen/Drives/Genoscope-googledrive/LabGroup_Shared/Lab_Projects/Flavobacterium/Cytometry/2026.04.15_Biostat_Wheatbran/FCS_files/";
file_in = "flavo_son_ble_F1_T1_1j__X4.fcs";
data_in = paste0(path_in, file_in);
data_1 = read.FCS(data_in, truncate_max_range = F);

```

# Main

Input data (data_1) is a flowFrame object with 3 slots

1.  exprs = matrix of events x channels
2.  parameters = AnnotatedDataFrame of meta data
3.  description = list of header info

```{r examine data}

# data summary
my_sum = summary(data_1);

# summary of measurements for each channel
total_events <- nrow(data_1)
cat("The total number of events acquired is:", total_events, "\n");

# summary of parameters
params = pData(parameters(data_1));
params <- as_tibble(params);
table = kable(params, caption = "Table: Acquired parameters. The range column describes the max value for each channel.");
table

```

Define the volume analyzed

```{r define volume analyzed}

kw = keyword(data_1);
vol_nl =  as.numeric(kw["$VOL"]);
vol_ml = vol_nl/1e6;

```

### Visualize data

Plot FSC vs SSC on linear scale

```{r plot on linear scale, warning = F, fig.cap = "Fig: FSC-A vs SSC-A on linear scale."}

# untransformed
plot_1 = autoplot(data_1, "FSC-A", "SSC-A") + 
         theme_minimal() + 
         labs(x = "FSC-A", y = "SSC-A");
plot_1

```

Plot on FSC vs SSC on log scale

```{r plot on log scale, fig.cap = "Fig: FSC-A vs SSC-A on log scale.", warning = F}

# log-transformed
plot_2 = autoplot(data_1, "FSC-A", "SSC-A") + 
    coord_cartesian(xlim = c(1, 100000), 
                    ylim = c(1, 100000))+
  scale_x_log10() + 
  scale_y_log10() + 
  labs(x = "FSC-A", y = "SSC-A")+
  theme_minimal();

plot_2

```

Plot FL1-A vs FL1-H to look for aggregates

```{r plot SYBR signal H vs A, fig.cap = "Fig: Plot of SYBR signal (FL1) area vs height. Single cells should appear on the diagononal, whereas cell aggregates are elsewhere."}

# log-transformed
plot_2a = autoplot(data_1, "FL1-H", "FL1-A") + 
    # coord_cartesian(xlim = c(1, 1e8), 
    #                 ylim = c(1, 1e8))+
  scale_x_log10() + 
  scale_y_log10() + 
  labs(x = "FL1-H", y = "FL1-A")+
  theme_minimal();

plot_2a

```

Univariate plot of SYBR fluorescence

```{r univariate log SYBR fluorescence, fig.cap = "Fig: Univariate plot of SYBR fluorescence on log scale.", warning = F}

plot_2b = autoplot(data_1, "FL1-A") + 
  scale_x_log10() + 
  theme_minimal()+
  labs(x = "FL1-A", y = "Density")+
  mytheme;
plot_2b

```

Logicle transform SYBR fluorescence

```{r logicle transform SYBR signal}

# Define the channels to transform
channels_to_transform <- c("FL1-H", "FL1-A");

# Estimate the logicle transformation parameters from your data
trans_list = estimateLogicle(data_1, channels_to_transform, m = 7);

# Apply the transformation to the flowFrame
data_1_transformed <- transform(data_1, trans_list);

```

Univariate density plot of logicle-transformed SYBR fluorescence

```{r time plot, fig.cap = "Fig: Logicle-transformed SSC_1-A signal."}

plot_3 = autoplot(data_1_transformed, "FL1-H") +   
  theme_minimal()+
  mytheme;
plot_3

```

Bivariate plot of logicle-transformed SYBR fluorescence

```{r bivariant SYBR fluorescence, fig.cap = "Fig: Bivariate plot of logicle transformed violet signal."}

# log-transformed
plot_4 = autoplot(data_1_transformed, "FL1-H", "FL1-A") + 
  theme_minimal();

plot_4

```

### Apply gates to filter data

Apply gate to remove lower peak of SYBR fluorescence

```{r violet gate, fig.cap = "Fig: plot of FL1-A data after gate to remove small events"}

# define gate bounds
gate_bounds <- list("FL1-A" = c(4.5, 5.5));

# define gate object
my_gate <- rectangleGate(filterId = "test gate", 
                         .gate = gate_bounds);

# apply gate
gate1_result <- filter(data_1_transformed, my_gate);
print(summary(gate1_result));

# subset gated data
data_gated_1 = Subset(data_1_transformed, my_gate);

# log-transformed
plot_5 = autoplot(data_gated_1, "FL1-A") +   
  theme_minimal()+
  mytheme;
plot_5

```

Replot gated FSC-A vs SSC-A

```{r replot gated data, warning = F, fig.cap = "Fig: FSC-A vs SSC-A after gate to remove small events."}

# log-transformed
plot_5b = autoplot(data_gated_1, "FSC-A", "SSC-A") + 
    coord_cartesian(xlim = c(10, 1000000), 
                    ylim = c(10, 1000000))+
  scale_x_log10() + 
  scale_y_log10() + 
  theme_minimal();

plot_5b


```

Apply rectangular gate to FSC-A, SSC-A

```{r gate FSC-A, fig.cap = "Fig: Rectanglular gate applied to FSC-A vs SSC-A after gate to remove small events.", warning = F}

# define gate bounds
gate_bounds <- list("FSC-A" = c(10, 50000), 
                    "SSC-A" = c(100, 100000))

# define gate object
my_gate <- rectangleGate(filterId = "rectangle gate", 
                         .gate = gate_bounds);

# apply gate
gate_result_r <- filter(data_1_transformed, my_gate);
print(summary(gate_result_r));

# subset data
data_gated_r = Subset(data_gated_1, my_gate);

# log-transformed
plot_5 = autoplot(data_gated_r, "FSC-A", "SSC-A") + 
    coord_cartesian(xlim = c(10, 1e6), 
                    ylim = c(10, 1e6))+
  scale_x_log10() + 
  scale_y_log10() + 
  theme_minimal();

plot_5


```

Apply polygon gate to FSC-A, SSC-A

```{r apply polygon gate, fig.cap = "Fig: Polygon gate applied to FSC-A vs SSC-A data.", eval = F}

# Define the coordinates for the polygon gate
polygon_coords <- matrix(c(100, 200, 
                           10, 5000, 
                           500, 70000,
                           30000, 70000,
                           30000, 200), 
                         ncol = 2, 
                         byrow = T,
                         dimnames = list(NULL, c("FSC-A", "SSC-A")));


# Create the gate object. 'polygonGate' converts the coordinates into an actual gated region 
# that flowCore can interpret.
poly_gate <- polygonGate(filterId = "MyPolygonGate", .gate = polygon_coords);

#xyplot(`SSC-A` ~ `FSC-A`, data = data_gated_1, filter = poly_gate);

# Apply the gate to the data. The 'filter' function checks every event 
# (cell) in 'data_1' to see if it falls inside the defined polygon.
gate_result_p <- filter(data_gated_1, poly_gate);
print(summary(gate_result_p));

# Extract the subset. 'Subset' creates a new object containing 
data_gated_p <- Subset(data_gated_1, gate_result_p);

# log-transformed
plot_6 = autoplot(data_gated_p, "FSC-A", "SSC-A") + 
    coord_cartesian(xlim = c(1, 70000), 
                    ylim = c(1, 70000))+
  scale_x_log10() + 
  scale_y_log10() + 
  theme_minimal();

plot_6

```

### Calculate the cell density (events / ml)

Calculate number of cells per ml

```{r calc cells per ml}

# 1. Get the number of events in your filtered data
n_events <- nrow(data_gated_r);

# Calculate concentration
concentration = dilution_factor * (n_events / vol_ml);
conc_sci = format(concentration, scientific = TRUE, digits = 3);

cat("The concentration is", conc_sci, "cells/mL\n")

```