---
title: "Quantify bacteria using Beckman Cytoflex"
author: "Andrew Tolonen"
format: html
editor: visual
toc: TRUE
code-fold: TRUE
embed-resources: TRUE
---

# Introduction

This file describes methods to visualize and filter flow cytometry data in .FCS files using the R flowCore package. Data was acquired using a Beckman CytoFlex instrument. Here, the samples are SYBR-stained bacterial cells. The goal is to accurately enumerate the number of bacteria in a defined volume.

## Background

FCS files contain all acquired data. However, running the Cytoflex requires 3 user-defined settings that affect what data is acquired:

-   Threshold = min signal for event, should set min signal manually.

-   Gain (hand signal on Cytoflex software). Note, the abort rate goes up if gain is too high.

-   Width = Set a width threshold, but other users stay with the default.

Also, the software asks the user to calculate number of events to save. This can be total events or total time. We defined a 20-30 sec of acquisition period with a flow rate of 10 ul per min.

### Description of channels

-   Forward scatter (488 nm): FSC-H, FSC-A, FSW-Width

-   Side scatter:

    -   blue laser (488 nm) = SSC-H, SSC-A

    -   violet laser (405 nm) = SSC_1-H, SSC_1-A (NOTE: Violet SSC provides enhanced resolution for small particles ie bacteria)

-   6 fluorescent channels: B525, B585, B690, R660, R780, DAPI.

    -   Letters are excitation. B = blue laser (488 nm). R = Red laser (638 nm). V = violet (405 nm)

    -   Numbers = emission wavelength

-   Time

# Setup

Libraries

```{r Libraries, message = F}

# libraries to include
library(knitr);
library(tidyverse);
library(readxl);
library(readr);
library(gridExtra);
library(grid);
library(ggpubr);
library(glue);
library(quarto);

library(flowCore);
library(ggcyto);

```

Flags/Config

```{r flags and config}

rm(list = ls());
setwd("/home/tolonen/Github/actolonen/Public/Lab_Public/Growth/Cytometry");

#Sys.setenv(QUARTO_PATH = "/home/tolonen/.conda/envs/Lab_Public/bin/quarto");
Sys.setenv(
  PATH = paste(
    "/home/tolonen/.conda/envs/Lab_Public/bin",
    Sys.getenv("PATH"),
    sep = .Platform$path.sep
  )
)

mytheme = theme(axis.text.x = element_text(size = 8), 
                axis.text.y = element_text(size = 8), 
                axis.title.x = element_text(size = 8), 
                axis.title.y = element_text(size = 8),
                strip.text.x = element_text(size = 8),
                strip.text.y = element_text(size = 8),
                legend.position = "none", 
                aspect.ratio = 0.25,
                plot.title = element_text(size = 8),
               );

# dilution factor of sample: cells are manually diluted before injection
dilution_factor = 4;

```

File IO

```{r File IO}

path_in = "/home/tolonen/Drives/Genoscope-googledrive/LabGroup_Shared/Lab_Projects/Flavobacterium/Cytometry/2026.04.15_Biostat_Wheatbran/FCS_files/";

file_in = "flavo_son_ble_F1_T1_1j__X4.fcs";

data_in = paste0(path_in, file_in);
data_1 = read.FCS(data_in, truncate_max_range = F, transformation = F);

```

# Main

## Input data

Input data (data_1) is a flowFrame object with 3 slots

1.  @exprs matrix of events x channels
2.  @parameters AnnotatedDataFrame of metadata with 4 slots
    1.  @data is a df with col "name" that lists channels: data_1\@parameters\@data\$name
3.  @description list of \~400 instrument settings

```{r examine data}

# data summary: distributions of data in each channel
my_sum = summary(data_1);

# summary of measurements for each channel
total_events_initial <- nrow(data_1)
output = paste0("The total number of acquired events: ", total_events_initial, "\n");
print(output);

# summary of channels
params = data_1@parameters;
channels = pData(params);
channels = channels %>%
    select(name, desc) %>%
    remove_rownames();

table = kable(channels, caption = "Table: Acquired channels. SYBR green (excite=500; emit=520) is captured by FL1. DAPI (excite=385, emit=488).");
table

```

Define the volume analyzed

```{r define volume analyzed}

# get header metadata from FCS file
kw = keyword(data_1);

# get volume of sample
vol_nl =  as.numeric(kw["$VOL"]);
vol_ul = vol_nl/1e3;
vol_ml = vol_nl/1e6;
output = paste0("The culture volume analyzed in this is sample is ", round(vol_ul, 2), " ul.")
print(output);

```

## Transform data

Logicle transform data

```{r logicle transform SYBR signal}

# Define the channels to transform
channels_to_transform <- c("FL1-H", "FL1-A", "FSC-A", "FSC-H", "SSC_1-A", "SSC_1-H");

# Estimate the logicle transformation parameters from your data
trans_list = estimateLogicle(data_1, channels_to_transform, m = 7);

# Apply the transformation to the flowFrame
data_transformed <- transform(data_1, trans_list);

```

## Gate data

### Gate 1: Margin/saturation exclusion

```{r saturation exclusion, fig.cap = "Fig: Density plots showing the distribution of values in the channels to analyze. There does not appear to be channel saturation. However, FSC-A has a wide distribution; might this be because the cells are rod-shaped?"}

# examine saturation using univariate plots

# extract data to plot as df
data_plot = as.data.frame(data_transformed@exprs);

# make list of plots for channels
plots.list = list();  # new empty list
counter = 1

for (i in seq_along(channels_to_transform)) {
  my_channel <- channels_to_transform[i]
  
  # Use .data[[my_channel]] to resolve lazy evaluation bug
  plots.list[[i]] <- ggplot(data_plot, aes(x = .data[[my_channel]])) +
    geom_density(fill = "skyblue", alpha = 0.5, color = "blue") +
    theme_classic() +
    labs(
      x = my_channel,               # Dynamic X-axis label
      y = "Density",
      title = my_channel # Dynamic title
    )+
    mytheme;
}

# Create the layout without auto-printing the object
plots.all <- grid.arrange(grobs = plots.list, ncol = 2, draw = FALSE)

# Explicitly draw only the graphics
grid.draw(plots.all)

# # Margin/saturation exclusion (remove events pinned at max range)
# range_gate <- rectangleGate("FSC-A" = c(0, 262143), 
#                            "SSC-A" = c(0, 262143), 
#                            "FITC-A" = c(0, 262143),
#                            filterId = "nonSaturated");

# # apply the gate
# data_saturation = Subset(data_transformed, range_gate);

data_gate_1 = data_transformed;

```

### Gate 2: SYBR fluorescence, side scatter

Plot: SSC (violet) vs SYBR fluorescence

```{r plot side-scatter vs SYBR fluorescence, warning = F, fig.cap = "Fig: Plot of SYBR fluorescence (x-axis) vs side scatter (y-axis). Red rectangle shows applied gate; percent is events passing the gate."}

# Define the gate boundaries on your transformed scale
my_gate <- rectangleGate(
  filterId = "SYBR_bacteria",
  "FL1-A"   = c(4.65, 5.7),  # min and max for X-axis
  "SSC_1-A" = c(3.9, 5)   # min and max for Y-axis
);

# Add the gate to autoplot
plot_sybr <- autoplot(data_transformed, "FL1-A", "SSC_1-A", bins = 1000) + 
  coord_cartesian(xlim = c(3, 8), ylim = c(3, 6)) +
  geom_gate(my_gate, colour = "red", size = 0.5) +  # Overlay gate line
  geom_stats(type = "percent") +                   # Optional: Display population %
  theme_classic();
plot_sybr

# apply the gate
data_gate_2 = Subset(data_gate_1, my_gate);

```

### Gate 3: double discrimination

Plot SSC-A vs SSC-H

```{r plot SSC-A vs SSC-H, fig.cap = "Fig: comparision of side scatter area (x-axis) vs height (y-axis)."}

# SSC
plot_doublet = autoplot(data_gate_2, "SSC_1-A", "SSC_1-H", bins = 500) + 
  coord_cartesian(xlim = c(3.5, 5.5), ylim = c(3.5, 5.5)) +
  labs(x = "SSC_1-A", y = "SSC_1-H") +
  theme_classic()
plot_doublet

```

Plot FSC-A vs FSC-H

```{r plot FSC-A vs FSC-H, fig.cap = "Fig: comparision of forward scatter area (x-axis) vs height (y-axis). This plot shows a wide range of areas for the same height, consistent with doublets."}

# FSC
plot_doublet = autoplot(data_gate_2, "FSC-A", "FSC-H", bins = 500) + 
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 5)) +
  labs(x = "FSC-A", y = "FSC-H") +
  theme_classic()
plot_doublet

```

Apply a polygon gate to the SSC-A vs SSC-H plot

```{r apply doublet gate}

# 1. Define vertex coordinates (order points sequentially around the shape)
gate_coords <- matrix(
  c(
    3.9, 3.85,  # Point 1 (x, y)
    5, 4.85,  # Point 2
    5, 5.1,  # Point 3
    3.9, 4.1),
  ncol = 2,
  byrow = TRUE,
  dimnames = list(NULL, c("SSC_1-A", "SSC_1-H")) # Must match object channel names
);

# 2. Create the polygonGate object
poly_gate <- polygonGate(filterId = "Singlets", .gate = gate_coords);

# Add the gate to autoplot
plot_doublet = autoplot(data_gate_2, "SSC_1-A", "SSC_1-H", bins = 500) + 
  coord_cartesian(xlim = c(3.5, 5.5), ylim = c(3.5, 5.5)) +
  labs(x = "SSC_1-A", y = "SSC_1-H") +
  geom_gate(poly_gate, colour = "red", size = 0.5) +  # Overlay gate line
  geom_stats(type = "percent") +                   # Optional: Display population %
  theme_classic()
plot_doublet

# apply the gate
data_gate_3 = Subset(data_gate_2, poly_gate);


```

### Gate 4: bacterial size

Plot SSC-A vs FSC-A

```{r plot SSC-A vs FSC-H, fig.cap = "Fig: comparision of forward scatter area (x-axis) vs side scatter area (y-axis)."}

# FSC
plot_size = autoplot(data_gate_3, "FSC-A", "SSC_1-A", bins = 500) + 
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 5)) +
  labs(x = "FSC-A", y = "SSC_1-A") +
  theme_classic()
plot_size

# apply gate 
data_gate_4 = data_gate_3;

```

### Calculate the cell density (events / ml)

Calculate number of cells per ml

```{r calc cells per ml}

# Get the number of events in your filtered data
total_events_final <- nrow(data_gate_4);
events_passing = round(100 * (total_events_final / total_events_initial), 2);
output = paste0(events_passing, "% of acquired events passed gating");
print(output);

# Calculate concentration
concentration = dilution_factor * (total_events_final / vol_ml);
conc_sci = format(concentration, scientific = TRUE, digits = 3);

output = paste0("The concentration is ", conc_sci, " cells/mL\n");
print(output);

```

# Conclusions

-   Univariate density plots show that FSC-A is a very wide distribution, which may be because the cells are rod-shared. I this focus on side-scatter for these analyses.

-   Violet side-scatter vs SYBR appears effective to differentiate cells from background.

-   SSC-A vs SSC-H appears effective to remove doublets.